---
title: "Project 3 - Cluster Analysis"
author: 
  - Jón Þorsteinsson - jth56@hi.is
  - Kristófer Már Gíslason - kmg14@hi.is
output: 
  html_document:
    css: styles.css
    theme: lumen
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cluster)
library(merTools)
library(dendextend)
library(colorspace) # get nice colors
```

### 1. Objectives

Our main obective was to determine with the highest possible accuracy what type of glass is found. Our ideal results would include 7 well defined clusters where each cluster would represent a glass type. The reason this can be found usefull is because of criminological investigations. Where at a scene of a crime glass can be correctly identified and possibly be used as valuable evidence.


### 2. Data set description

Without any preprocessing our data set has 214 different instances that each contain 10 different attributes and one class attribute that tells us what kind of glass it is from the 7 possible options.

| Nr. | Name | Type | Description |
| --- | ---- | ---- | ----------- |
| 1. | Id| numerical | number: 1 to 214 |
| 2. | RI| numerical | Refractive Index |
| 3. | Na| numerical | Sodium |
| 4. | Mg| numerical | Magnesium |
| 5. | Al| numerical | Aluminum |
| 6. | Si| numerical | Silicon |
| 7. | K | numerical | Potassium |
| 8. | Ca| numerical | Calcium|
| 9. | Ba |numerical | Barium |
| 10. |Fe |numerical | Iron| 
| 11. | Class | nominal | class - (building_windows_float_processed,building_windows_non_float_processed,vehicle_windows_float_processed, vehicle_windows_non_float_processed,containers,tableware,headlamps)|

### 3. Preprocessing

We began by visualizing the data and coloring each point according to class. We can see that all the classes seem to be pretty spread out over the plots and there doesn't seem to be any particular pattern that could help us with the clustering but there are however a few obvious outliers. We remove them by hand and then the data looks a little bit better. At this stage we try looking at all our attributes and noticed that the first one **Id** is completely useless to us because it only numbers the instances which we don't need. We also took out the class variable so we can use that to evaluate our results. We also noticed that the class attributes never results to the glass being *vehicle_windows_non_float_processed* for this dataset so we should get one fewer clusters in the final result. All other attributes seem like they might have some interesting information so we include them.
It is also imperative to normalize the scale of feature values in order to begin with the clustering process. This is because each observations' feature values are represented as coordinates in n-dimensional space


```{r data, tidy = TRUE}
# Reading in data file and viewing our data
glass.full <- read.csv("glass.csv")
plot(glass.full, col=glass.full$class)

# Removing some obvious outliers by hand
glass.full <- glass.full[-c(172, 173, 202, 107, 164, 208, 185, 175, 108, 186, 187),]
plot(glass.full, col=glass.full$class)

# Removing the Id and Class attribute
glass <- glass.full[2:10]

# Normalizing all attributes 
glass <- scale(glass)
```

### 4. K-means clustering

Now that we have finished preprocessing the next step is to run the k-means algorithm but first we need to find what number of clusters that will give us the best result we accomplish this by plotting the sum of squares and seeing when the drop decreases. Since the initial cluster assignments are random, we need to set the seed to ensure reproducibility. 
By plotting the sum of squares with a few different seed values we can see that the drop of the y-axis decreases when the number of clusters is around 5 and 6 this is makes sense since the number of classes we have is 6.

```{r kmeans}
# SEED = 12
set.seed(12)
# Determine number of clusters
wss1 <- (nrow(glass)-2)*sum(apply(glass,2,var))
for (i in 2:15) wss1[i] <- sum(kmeans(glass,centers=i)$withinss)

# SEED = 23
set.seed(23)
# Determine number of clusters
wss2 <- (nrow(glass)-2)*sum(apply(glass,2,var))
for (i in 2:15) wss2[i] <- sum(kmeans(glass,centers=i)$withinss)

# SEED = 34
set.seed(34)
# Determine number of clusters
wss3 <- (nrow(glass)-2)*sum(apply(glass,2,var))
for (i in 2:15) wss3[i] <- sum(kmeans(glass,centers=i)$withinss)

# SEED = 45
set.seed(45)
# Determine number of clusters
wss4 <- (nrow(glass)-2)*sum(apply(glass,2,var))
for (i in 2:15) wss4[i] <- sum(kmeans(glass,centers=i)$withinss)

# SEED = 56
set.seed(56)
# Determine number of clusters
wss5 <- (nrow(glass)-2)*sum(apply(glass,2,var))
for (i in 2:15) wss5[i] <- sum(kmeans(glass,centers=i)$withinss)

# Plotting the graph
plot(1:15, wss1, type="b", xlab="Number of Clusters", ylab="sum of squares")
points(wss2, col='red', )
points(wss3, col='blue')
points(wss4, col='green')
points(wss5, col='orange')
```


```{r}
set.seed(12)

# running k-means 
glassClusterKmeans <- kmeans(glass, 6, nstart = 20, iter.max=500)

# Printing a table showing how the class labels fit compared to the clusters. 
table(glassClusterKmeans$cluster, glass.full$class)

# reviewing the total sum of squared errors  
glass_kmeans_error <- sum(glassClusterKmeans$withinss)
glass_kmeans_error

# The plot of our data with clusters represented with different colors.
plot(glass.full, col=glassClusterKmeans$cluster,)
```
Like we expected after looking at our data the kmeans clustering did not give us very good results. The clusters are mostly grouped together in bigger clusters and it's very hard to see any distinction between them in any of the plots even when we look at them individually. **Reikna entropy og purity**

### 5. Hierarchical Clustering
Next we use *hclust* to calculate and plot the Hierarchical Clustering with the already preproccessed data. *hclust* requires us to provide the data in the form of a distance matrix. We can do this by using *dist*. We decide too try both average and ward and to see which gives us a better result.

```{r hiearchy}

fit <- dist(glass, method = 'euclidean')
d.ward <- hclust(fit, method = "ward.D")
d.average <- hclust(fit, method = "average")

# Create a dendrogram object from the hclust variable
dend.ward <- as.dendrogram(d.ward)
dend.average <- as.dendrogram(d.average)

# Color brances by cluster formed from the cut at a height of 6
colDat.ward <- color_branches(as.dendrogram(d.ward), h=6)
colDat.average <- color_branches(as.dendrogram(d.average), h = 6)

# Plot the dendrogram 
plot(dend.ward)
plot(dend.average)
```

The average method gives us a terrable result se we can just throw that away. When we look at the tree for the ward method we can see that a cut at height 6 could work well which makes sense because of our number of classes.

```{r hiearchy_2}
#WARD
# distance matrix
d <- dist(glass, method = "euclidean") 
fit.ward <- hclust(d, method = "ward.D")

# display dendogram
plot(fit.ward)

# cut tree into 6 clusters
groups.ward <- cutree(fit.ward, k=6)

# draw dendogram with blue borders around the 6 clusters 
rect.hclust(fit.ward, k=6, border="blue")

table(groups.ward, glass.full$class)
```


The ouput after cutting the tree is still not very efficient we can't really tell any results from it. From the table we can see that most of the instances are put in to the first two clusters and the distribution in those clusters is pretty even between 2 to 3 classes. **Reikna entropy og purity**


### 6. Comparison of K-means and Hierarchical

Before we compare the k-means and Hierarchial clustering we must acknowledge what differences thes two clustering methods have. Both have flaws and strengths such as the Hierarchical clustering can virtually handle any distance metric while k-means relys on euclidean distances. k-means doesn't have the same stability of results as Hierarchial because k-means requires a random step at its initialization that may yield different results if the process is re-run. That wouldn't be the case in hierarchical clustering.
K-means is also less computationally expensive than hierarchical clustering and can be run on large datasets within a reasonable time frame, which is the main reason k-means is more popular. 
The results we got indicated that neither clustering algorithm was suitable for our data which might then in return imply that our data is just not suited for clustering










